On discrete entropy conservation for general high order discontinuous Galerkin methods

### 2017-07-13 docs/splitform/main.tex
updating with new paper

\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
%\usepackage{thmtools}
%\declaretheorem[style=definition,qed=$\blacksquare$,numberwithin=chapter]{definition}

@@

\date{}
\author{Jesse Chan}
\title{Energy stable discontinuous Galerkin formulations using discrete differential operators}

@@

\begin{abstract}
We show that, for non-linear hyperbolic conservation laws which admit a skew-symmetric splitting, energy stable discontinuous Galerkin (DG) methods can be constructed in a straightforward manner based on discrete DG gradient and divergence operators.  In particular, we show that these formulations remain energy stable for both curvilinear geometries and inexact quadrature.  Examples of energy stable formulations are given for variable advection, Burgers' equation, and a Burgers'-like system, and it is shown that the construction of DG methods based on discrete differential operators recover known entropy-conservative fluxes.  
\end{abstract}

@@

Non-conservative discretizations \cite{rhebergen2008discontinuous}.

DG calculus \cite{feng2016discontinuous} usually assumes exact integration.  As pointed out in \cite{gassner2016split}, this can be either be very expensive (curvilinear meshes or polynomial nonlinearities) or impossible (in the presence of rational integrands, such as for compressible flow).

Lifting operators and discrete gradients \cite{bassi1997high, di2011mathematical}.

Goal: reduce aliasing in advective problems caused by variable coefficients and curvilinear or non-affine mappings.  Resembles ideas from summation by parts \cite{olsson1995summation1, olsson1995summation2, hicken2016multidimensional}, but based on global DG differentiation operators.

\section{Discrete differential operators}

We assume that the domain $\Omega$ is decomposed into non-overlapping elements $D^k$.  We define the mesh $\Omega_h = \cup D^k$ and the corresponding global approximation space $V_h(\Omega_h) = \bigoplus V_h\LRp{D^k}$, where $V_h\LRp{D^k}$ is the approximation space over a single patch.  Furthermore, we introduce the jump and average of discontinuous functions across element interfaces.  Let $D^{k,+}$ denote the neighboring element of across a face $f$ of $D^k$, and let $u^+,u^-$ denote the values of $u$ on $D^{k,+}$ and $D^k$, respectively.  The jump of $u$ across $f$ is then defined as
\[
\jump{u} = u^+ - u^-, \qquad \avg{u} = \frac{u^+ + u^-}{2}.
\]
On faces which coincide with the boundary $\partial \Omega$, the average and jump are defined as
\[
\jump{u} = 0, \qquad \avg{u} = u.
\]
The jump and average of vector fields are defined component-wise using the jumps and averages of components.

We will assume for now that all elements are affine mappings of a reference element $\widehat{D}$.  Extensions to curvilinear meshes and non-affine mappings will be discussed in Section~\ref{sec:curv}.

\subsection{Discrete inner products}

In order to define discrete differential operators, we first introduce the $L^2$ inner product on $V_h\LRp{\Oh}$
\[
\LRp{u,v}_{\Omega} = \sum_{k} \int_{D^k} uv \diff{x} = \sum_{k} \int_{\widehat{D}} uv J \diff{\widehat{x}},
\]
and the associated $L^2$ projection $\Pi_N: L^2\LRp{\Omega}\rightarrow P^N$
\[
\LRp{\Pi_N u,v}_{\Omega} = \LRp{u,v}_{\Omega}.  
\]

In practice, these integrals are computed using quadrature, such that 
\begin{equation}
\int_{\widehat{D}}uv = \sum_{i=1}^{N_q} u(\bm{x}_i)v(\bm{x}_i) w_i,
\label{eq:quad}
\end{equation}
where $N_q$ is the number of quadrature points.  The only assumptions we make upon this quadrature is that it is sufficiently accurate such that 
\begin{enumerate}
\item The quadrature induces an $L^2$-equivalent inner product over the reference element $\widehat{D}$.
\item The quadrature is sufficiently accurate such that integration by parts holds with respect to the reference coordinates $\widehat{x}$.
\end{enumerate}

\subsection{Discrete derivatives}
We introduce two discrete DG derivatives in this section, based on definitions used in \cite{di2011mathematical} and \cite{hesthaven2004high,Warburton20063205}:
\begin{definition}
We define two discontinuous Galerkin differentiation operators ${D}^i: L^2\LRp{\Omega}\rightarrow V_h$ and $\tilde{D}^i: L^2\LRp{\Omega}\rightarrow V_h$ is defined implicitly as follows:
\begin{align}
\LRp{{D}^i u,v}_{\Oh} &= \sum_{k} \LRp{\LRp{-u,\pd{v}{\bm{x}_i}}_{\LK} + \LRa{\avg{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v \in V_h\\
\LRp{\tilde{D}^i u,v}_{\Oh} &= \sum_{k} \LRp{\LRp{\pd{u}{\bm{x}_i},v}_{\LK} - \frac{1}{2}\LRa{\jump{u},v\bm{n}_i}_{\LdK}}, \qquad \forall v\in V_h.
\label{eq:dgd1}
\end{align}
\end{definition}

We note that, assuming $u,v \in V_h$ and sufficiently accurate quadrature, $D^i$ and $\tilde{D}^i$ are equivalent.  The main property of the discrete derivatives utilized in this work is a discrete integration-by-parts formula.  
\begin{lemma}
Let $u,v\in V_h$.  Then, 
\begin{align*}
\LRp{D^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
%\LRp{D_h^i u,v} &= \LRp{-u,D^i_h v} + \int_{\partial \Omega} uv \bm{n}_i.
\end{align*}
\label{lemma:ibp}
\end{lemma}
\begin{proof}
Assuming that quadrature is sufficiently accurate for $u, v\in V_h$, integration by parts gives the equivalence. 
\end{proof}

For $u \not\in V_h$, due to the fact that the quadrature rule (\ref{eq:quad}) may not be exact, and these two definitions are not equivalent and Lemma~\ref{lemma:ibp} does not hold.  This motivates the definition of a \textit{discrete} DG differentiation operator:
\begin{definition}
The discrete discontinuous Galerkin differentiation operator ${D}^i_h: L^2\LRp{\Omega} \rightarrow V_h$ is defined as $D^i_h = D^i \Pi_N$.
\end{definition}

Using these discrete derivatives, we can define discrete gradient and divergence operators
\begin{definition}
Let $u \in V_h$ and $\bm{u}\in \LRp{V_h}^d$.  Then, discrete gradient and divergence operators are defined as
\begin{align*}
\LRp{\Grad^{\rm DG} u}_i &= D^i u, \qquad \Grad^{\rm DG}\cdot \bm{u} = \sum_{i=1}^d D^i \bm{u}_i\\
\LRp{\Grad^{\rm DG}_h u}_i &= D_h^i u, \qquad \Grad^{\rm DG}_h\cdot \bm{u} = \sum_{i=1}^d D_h^i \bm{u}_i.
\end{align*}
\end{definition}

When the support of $v$ is limited to a single element, we have
\[
\LRp{\Grad_h\cdot \bm{u},v\bbm{1}_{D^k}} = \LRp{\bm{u}, \Grad\cdot v\bbm{1}_{D^k}}  + \LRa{\avg{\bm{u}}\cdot\bm{n},v}_{\partial D^k}.
\]
and as a result when $v = 1$
\[
\LRp{\Grad_h\cdot \bm{u},\bbm{1}_{D^k}} = \int_{\partial D^k}\avg{\bm{u}}\cdot\bm{n}
\]

\section{Examples}

\begin{itemize}
\item Tensor product elements: inner product computed using GLL nodes, $\Pi_N$ reduces to nodal interpolation
\item General elements: $\Pi_N$ is just quadrature-based projection.  For accuracy, degree $2N$ or higher?  
\item Triangles and tetrahedra: lumped quadrature nodes.  Implies that mass lumping on triangles requires sufficiently accurate quadrature such that IBP holds.  This conclusion was first reached in the study of multi-dimensional SBP operators \cite{hicken2016multidimensional}.  
\item General elements: $\Pi_N$ can be taken to be nodal interpolation.  
\item Pyramids: need to account for non-affine mappings.  
\end{itemize}

\subsection{Variable advection}

A split formulation for advection is 
\[
\LRp{\pd{u}{t},v} + \frac{1}{2}\LRp{\Grad_h\cdot \Pi_N \LRp{ \bm{\beta}u},v} + \frac{1}{2}\LRp{\bm{\beta}\cdot\Grad_h u,v} + \frac{1}{2}\LRp{\LRp{\Grad\cdot \bm{\beta}} u,v} = 0.
\]
Taking $v = u$ yields and using $\LRp{\Grad_h u, \bm{v}} = \LRp{-u, \Grad_h \cdot \bm{v}}$ yields the energy statement
\[
\frac{1}{2}\nor{u}^2 + \frac{1}{2}\LRp{\Grad_h\cdot \Pi_N \LRp{ \bm{\beta}u},u} - \frac{1}{2}\LRp{ u, \Grad_h\cdot \Pi_N\LRp{\bm{\beta}u}} = \frac{1}{2}\LRp{-\LRp{\Grad \cdot \bm{\beta}} u,u},
\]
implying that $\frac{1}{2}\nor{u}^2 = 0$ if $\Grad\cdot \bm{\beta} = 0$, or that the method is energy conserving.  The only difference in this formulation is the introduction of $\Pi_N$, which can be defined at a discrete level using any quadrature scheme for which a discrete projection is well-defined.

Penalization can be added by adding any positive-definite stabilization term (upwind, penalty, Lax-Friedrichs) through the regular divergence flux.

\section{Discrete DG derivatives}

Methods based on discrete DG derivatives also work.

The discrete DG derivative-based method is not consistent in the sense that Galerkin orthogonality does not hold exactly.  The difference lies in the flux terms.  Assume $\Div \bm{\beta} = 0$, then
\[
\LRp{\pd{u}{t},v} + \frac{1}{2}\LRp{-{ \bm{\beta}u},\Grad_h v} + \frac{1}{2}\LRp{\bm{\beta}\cdot\Grad_h u,v} = 0.
\]
\begin{align*}
\LRp{\Grad_h\cdot \Pi_N\LRp{ \bm{\beta}u},v} &= \sum_{k} \LRp{-\bm{\beta}u, \Grad v}_{\LK} + \LRa{\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n},v}_{\LdK}\\
\LRp{\Grad_h u,\bm{\beta}v} &= \sum_{k} \LRp{-u, \Grad \cdot \LRp{\bm{\beta}v}}_{\LK} + \LRa{\bm{\beta}\cdot\bm{n}\avg{u},v}_{\LdK}.
\end{align*}
The latter term is consistent; the former is not due to the presence of $\avg{\Pi_N\LRp{\bm{\beta} u}}\cdot \bm{n}$ in the flux term.  The consistency error should then be $O(h^{N+1/2})$ using a trace inequality for $L^2$ projections.

Note: can also use interpolants in a stable manner if using $D_h$.  Unlike SEM, this still requires an extra matvec per RHS evaluation because of the lack of diagonality of the mass matrix.  Reduces number of steps by one (no interpolation to quadrature points) but does not reduce number of total matvecs.

\subsection{Local conservation}

Writing this in non-conservative form raises the question of local conservation.  Integrating the original equation over $D^k$ and using Gauss' theorem gives
\[
\int_{D^k}\pd{u}{t} + \int_{\partial D^k} \beta_n u = 0.
\]
Taking $v = 1$ on $D^k$ yields
\[
\int_{D^k}\pd{u}{t} + \frac{1}{2}\LRp{\Grad_h\cdot \Pi_N \LRp{ \bm{\beta}u},\bbm{1}_{D^k}} + \frac{1}{2}\LRp{\bm{\beta}\cdot\Grad_h u,\bbm{1}_{D^k}} + \frac{1}{2}\LRp{\LRp{\Grad\cdot \bm{\beta}} u,\bbm{1}_{D^k}} = 0.
\]
The first term gives
\[
\LRp{\Grad_h\cdot \Pi_N \LRp{ \bm{\beta}u},\bbm{1}_{D^k}} = \int_{\partial D^k} \avg{\Pi_N\LRp{\bm{\beta}u}}\cdot\bm{n}.
\]
The second term gives
\[
\LRp{\bm{\beta}\cdot\Grad_h u,\bbm{1}_{D^k}} = \LRp{\Grad u,\bm{\beta}}_{D^k} + \frac{1}{2}\LRa{\jump{u},\bm{\beta}\cdot\bm{n}} = \LRp{u,-\Grad \cdot \bm{\beta}}_{D^k} + \LRa{\avg{u},\bm{\beta}\cdot\bm{n}}
\]
through integration by parts and an assumption that $\bm{\beta}\cdot \bm{n}$ is periodic.  Cancelling volume terms, we end up with the statement of local conservation 
\[
\int_{D^k} \pd{u}{t}  + \frac{1}{2}\int_{\partial D^k} \LRp{\avg{\Pi_N\LRp{\bm{\beta}u}} + \LRp{\Pi_N\LRp{\bm{\beta}}\avg{u}}}\cdot\bm{n} = 0
\]
which is a discrete version of the continuous statement of local conservation.

\section{Non-affine mappings}
\label{sec:curv}

For general elements and curvilinear meshes, $J$ and $\bm{G}$ are no longer constant over each element.  However, for isoparametric mappings (and appropriate polynomial interpolations of general curvilinear mappings), metric identities hold \cite{kopriva2006metric} such that $\widehat{\Grad} \cdot \LRp{J\bm{G}^T} = 0$.  Thus, we can write
\[
\LRp{\Grad u,\bm{v}}_{D^k} = \LRp{J\bm{G}\widehat{\Grad} u,\bm{v}}_{\widehat{D}} = \LRp{-u,\widehat{\Grad} \cdot \LRp{J\bm{G}\bm{v}}}  + \LRa{J\bm{G}^T\widehat{\bm{n}} u,\bm{v}}.   
\]
Noting that $J\bm{G}\widehat{\bm{n}} = J^f \bm{n} $ \cite{hesthaven2007nodal}, we recover integration by parts over the physical element.

Thus, to extend the discrete discontinuous Galerkin gradient and divergence to non-affine mappings, we simply define
\begin{align*}
\LRp{\widehat{\Grad}_h u,\bm{v}}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}u,\bm{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{u},\bm{v}} \\
\LRp{\widehat{\Grad}_h \cdot\bm{u},{v}}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}\cdot \bm{u},{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\jump{\bm{u}},v\widehat{\bm{n}}} 
%\LRp{\Grad_h \cdot \bm{u},v}_{\Omega} &= \sum_{k} \LRp{\widehat{\Grad}\cdot{\Pi_N\LRp{J\bm{G}\bm{u}}},v}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{J\bm{G}\bm{u}},v} \\
%\LRp{\Grad_h u,\bm{v}}_{\Omega} &= \sum_{k} \LRp{\Pi_N\LRp{J\bm{G}\widehat{\Grad} u},\bm{v}}  - \frac{1}{2}\LRa{\widehat{\bm{n}} \jump{J\bm{G}u},v}.
\end{align*}
and define the discrete physical gradient and divergence as
\begin{align*}
\LRp{\Grad_h u,v}_{\Omega} &= \LRp{J\bm{G}\widehat{\Grad}_h u,\bm{v}}_{\Omega} = \sum_{k}\LRp{J\bm{G}\widehat{\Grad}u,\bm{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{J\bm{G}\widehat{\bm{n}} \jump{u},\bm{v}} \\
\LRp{\Grad_h \cdot \bm{u},v}_{\Omega} &= \LRp{\widehat{\Grad}_h\cdot\LRp{J\bm{G}^T\bm{u}}}_{\Omega} = \sum_{k} \LRp{\widehat{\Grad}\cdot \LRp{J\bm{G}^T\bm{u}},{v}}_{\widehat{D}}  - \frac{1}{2}\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}}. 
\end{align*}
The discrete gradient is consistent with the physical gradient.  The discrete divergence is also consistent with the physical divergence, provided that the jump term $\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}}$ is computed via
\[
\LRa{\jump{J\bm{G}^T\bm{u}},v\widehat{\bm{n}}} = \LRa{\jump{J\LRp{\bm{G}\widehat{\bm{n}}}^T\bm{u}},v} = \LRa{\LRp{\widehat{\bm{n}}^T\LRp{J{\bm{G}}^T\bm{u}}}^+ + \LRp{\widehat{\bm{n}}^T\LRp{J{\bm{G}}^T\bm{u}}}^-,v}.  
\]

The advantage of this projected geometric factor approach is local conservation; taking $v = \bbm{1}_{D^k}$ recovers (after integration by parts) the flux integral
\[
\LRp{\Grad_h \cdot \bm{u},\bbm{1}_{D^k}}_{\Omega} = \int_{\partial D^k}\widehat{\bm{n}}^T\avg{J\bm{G}^T\bm{u}}.
\]

\section{Extension to other hyperbolic problems}

>>>>

\usepackage{fullpage}
\usepackage{amsmath,amssymb,amsfonts,amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem*{remark}{Remark}
\theoremstyle{theorem}
\newtheorem{theorem}{Theorem}
%\usepackage{thmtools}
%\declaretheorem[style=definition,qed=$\blacksquare$,numberwithin=chapter]{definition}

@@

\date{}
\author{Jesse Chan}
\title{On discrete entropy conservation for general high order discontinuous Galerkin methods}

@@

\begin{abstract}
High order methods based on diagonal-norm summation by parts operators can be shown to semi-discretely conserve entropy for nonlinear systems of hyperbolic PDEs \cite{fisher2013high,carpenter2014entropy}.  These include discontinuous Galerkin spectral element methods (DG-SEM) \cite{gassner2016split,gassner2016well,wintermeyer2017entropy, chen2017entropy} based on mass-lumping.  In this work, we describe how to extend discretely entropy conservative schemes for a more general class of DG methods using quadrature-based projections and global differentiation operators.  The construct of these methods is independent of the choice of basis and accuracy of quadrature rule used, and includes DG methods involving generalized SBP operators without boundary nodes and certain dense-norm SBP operators.  Numerical experiments confirm the stability and high order accuracy of the proposed methods for the Burgers, shallow water, and compressible Euler equations.  
\end{abstract}

\tableofcontents

@@

\note{High order methods are good boilerplate}

\note{High order methods are not stable for nonlinear problems.  Stability of discretizations is a hot topic: artificial viscosity, polynomial de-aliasing,  \cite{karniadakis2013spectral}}

\cite{tadmor1987numerical, tadmor2003entropy}

Hughes entropy variables \cite{hughes1986new}

\cite{hicken2016multidimensional}

\cite{gassner2016split,gassner2016well,wintermeyer2017entropy}

\cite{chen2017entropy}

While DG-SEM based on flux differencing is entropy conservative, this is tied to a specific choice of basis and quadrature.  Furthermore, GLL quadratures and diagonal-norm SBP operators do not naturally generalize to triangles or tetrahedra \cite{helenbrook2009existence}.  Diagonal-norm SBP operators can be constructed for triangles and tetrahedra \cite{chin1999higher, cohen2001higher, hicken2016multidimensional, chen2017entropy}; however, the number of nodal points for such operators is greater than the dimension of the natural polynomial approximation space.  Furthermore, appropriate point sets have only been constructed for $N \leq 4$ in three dimensions \cite{zhebel2014comparison}.

\section{Entropy conservative collocation DG methods}

\subsection{Entropy stability for systems of hyperbolic PDEs}

We consider systems of nonlinear conservation laws in $\mathbb{R}^d$ with $n$ variables
\begin{align}
\pd{\bm{u}}{t} + \sum_{k=1}^d \pd{\bm{f_k(\bm{u})}}{\bm{x}_k} &= 0, \qquad \bm{u}(\bm{x},t) = (u_1(\bm{x},t),\ldots,u_n(\bm{x},t)).
\label{eq:pde}
\end{align}
where the fluxes $f_k(\bm{u})$ are smooth functions of the vector of conservative variables $\bm{u}(\bm{x},t)$.
The Jacobian matrix $\bm{A}_k(\bm{u})$ is defined entrywise as
\[
\LRp{\bm{A}_k(\bm{u})}_{ij} = \LRp{\pd{\bm{f}_k(\bm{u})}{\bm{u}_j}}_i.
\]
We are interested in systems for which there exists a convex entropy function $U(\bm{u})$ such that  
\begin{equation}
U''(\bm{u})\bm{A}_k(\bm{u}) = \LRp{U''(\bm{u}) \bm{A}_k(\bm{u})}^T.
\label{eq:entropysym}
\end{equation}
For systems with convex entropy functions, one can define entropy variables $\bm{v} = U'(\bm{u})$.  The convexity of the $U(\bm{u})$ guarantees that the mapping between conservative and entropy variables is invertible.

It can be shown (see, for example, \cite{mock1980systems}) that (\ref{eq:entropysym}) is equivalent to the existence of an entropy flux functions $F_k(\bm{u})$ such that
\[
\bm{v}^T \pd{\bm{f}_k}{\bm{u}} = \pd{F_k(\bm{u})}{\bm{u}}^T.
\]
The flux function and entropy flux function are further related through the so-called entropy potential $\psi_k$ 
\[
\psi_k(\bm{v}) = \bm{v}^T\bm{f}_k(\bm{u}(\bm{v})) - F_k(\bm{u}(\bm{v})), \qquad \psi_k'(\bm{v}) = \bm{f}_k(\bm{u}(\bm{v})).
\]

When $\bm{u}$ is smooth, multiplying (\ref{eq:pde}) on the left by $\bm{v}^T = U'(\bm{u})^T$, applying the definition of the entropy flux and using the chain rule yields the conservation of entropy
\[
\pd{U(\bm{u})}{t} + \sum_{k=1}^d \pd{F_k(\bm{u})}{\bm{x}_k} = 0.
\]
More generally, it can be shown that physically relevant solutions of (\ref{eq:pde}) (defined as the limiting solution for an appropriately defined vanishing viscosity) satisfy the entropy inequality
\[
\pd{U(\bm{u})}{t} + \sum_{k=1}^d \pd{F_k(\bm{u})}{\bm{x}_k} \leq 0.  
\]
%Integration over $\mathbb{R}^d$ then shows that the entropy of the solution decreases in time.  
When $d = 1$, integrating the entropy inequality over $[-1,1]$ and using the definition of the entropy potential yields
\begin{equation}
\int_{-1}^1 \pd{U(\bm{u})}{t} + \left.\LRp{\bm{v}^T\bm{f}_k(\bm{u}(\bm{v}))-\psi(\bm{v})}\right|_{-1}^1 \leq 0.  
\label{eq:consentropy}
\end{equation}

%\note{Wave equation example}

\subsection{Entropy conservative collocation methods based on SBP operators}

Let $\widehat{x}_i \in [-1,1]$ denote the $(N+1)$ point Gauss-Lobatto-Legendre (GLL) quadrature rule with positive weights $w_1,\ldots,w_{N+1}$.  Let $\ell_i(x)$ denote the nodal Lagrange basis defined at GLL points, and define the differentiation matrix $\bm{D}$ be defined as 
\[
\bm{D}_{ij} = \pd{\ell_j(x_i)}{x}.
\]
The matrix $\bm{D}$ maps from nodal values of a polynomial to nodal values of its derivative.  We also define the mass matrix $\bm{M}$
\[
\bm{M}_{ij} = \int_{-1}^1 \ell_i(x)\ell_j(x)\diff{x} \approx \delta_{ij} w_i,
\]
where the approximation results evaluating the integral using GLL quadrature.  For this choice of quadrature, the Kronecker property $\ell_j(x_i) = \delta_{ij}$ gives that $\bm{M}$ is a diagonal matrix.

The mass and derivative matrices may be used to define the diagonal norm SBP operator $\bm{S} = \bm{M}\bm{D}$, with the summation-by-parts property 
\[
\bm{S} = \bm{B} - \bm{S}^T, \qquad \bm{B}_{ij} = \begin{cases}
1 &i = j = 1 \\
1 &i = j = N+1 \\
0 &\text{otherwise}
\end{cases}.
\]
This property is simply a restatement of integration by parts, using that the derivative of $\ell_j(x) \in P^{N-1}$ and that GLL quadrature is exact for polynomials of degree $2N-1$.

%In order to simplify notation for flux differencing, we define the following bilinear matrix product 
%\[
%\LRp{\bm{A} \odot \bm{f}_S(\bm{x},\bm{y})}_i = \sum_j \bm{A}_{ij} \bm{f}_S(\bm{x}_j,\bm{y}_i).  
%\]
We seek a nodal approximation $\bm{u}_i \approx \bm{u}(x_i)$ to the solution $\bm{u}(x)$ of (\ref{eq:pde}).  A typical nodal collocation method in conservation form is given as follows 
\[
\pd{\bm{u}_i}{t} + \sum_{j=1}^{N+1}\LRp{\LRp{\bm{D}\otimes \bm{I}}_{ij}\bm{f}(\bm{u}_j) +\LRp{\bm{M}^{-1}\bm{B}}_{ij}(\bm{f}_i^* - \bm{f}(\bm{u}_i))} = 0.
\]
Here, $\bm{f}^*$ is a numerical flux used for the weak imposition of boundary or continuity conditions \cite{hesthaven2007nodal}.  This method can be derived from an appropriate DG formulation under a specific choice of quadrature.

This formulation conserves the quantities $\bm{u}_i$, but (except in specific cases) does not satisfy a discrete analogue of the statement of entropy conservation (\ref{eq:consentropy}).  Entropy conservative schemes may be constructed based on the SBP property and a ``flux differencing'' approach \cite{gassner2016split} involving a definition of entropy conservative/stable fluxes given by Tadmor \cite{tadmor1987numerical, tadmor2003entropy}.  
\begin{definition}
Let $\bm{f}_S(\bm{u}_L,\bm{u}_R)$ be a bivariate function such that it is symmetric and consistent with the flux function $\bm{f}(\bm{u})$
\[
\bm{f}_S(\bm{u}_L,\bm{u}_R) = \bm{f}_S(\bm{u}_R,\bm{u}_L), \qquad \bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u})
\]
The numerical flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is entropy conservative if, for entropy variables $\bm{v}_L = \bm{v}(\bm{u}_L), \bm{v}_R = \bm{v}(\bm{u}_R)$
\[
\LRp{\bm{v}_L - \bm{v}_R}^T \bm{f}_S(\bm{u}_L,\bm{u}_R) = (\psi_L - \psi_R), \qquad \psi_L = \psi(\bm{v}(\bm{u}_L)), \quad \psi_R = \psi(\bm{v}(\bm{u}_R)).  
\]
Similarly, a flux $\bm{f}_S(\bm{u}_L, \bm{u}_R)$ is referred to as entropy stable if $\LRp{\bm{v}_L - \bm{v}_R}^T \bm{f}_S(\bm{u}_L,\bm{u}_R) \leq (\psi_L - \psi_R)$.
\label{def:tadmor}
\end{definition}
An entropy conservative nodal collocation method can then be derived by modifying the flux derivative as follows
\begin{equation}
\pd{\bm{u}_i}{t} + \sum_{j=1}^{N+1}\LRp{\LRp{2\bm{D}\otimes \bm{I}}_{ij}\bm{f}_S(\bm{u}_i,\bm{u}_j) +\LRp{\bm{M}^{-1}\bm{B}}_{ij}(\bm{f}_i^* - \bm{f}(\bm{u}_i))}= 0.
\label{eq:dgsem}
\end{equation}
The replacement of the derivative of the flux function with the flux differencing allows one to prove that (\ref{eq:dgsem}) conserves a discrete entropy \cite{gassner2017br1,chen2017entropy}.  We reproduce these proofs here, as we will refer to them in following sections.  
\begin{theorem}
The collocation method (\ref{eq:dgsem}) is discretely entropy conservative in the sense that
\[
\bm{1}^T\LRp{\bm{M}\pd{U(\bm{u})}{t} + \bm{B}\LRp{\bm{v}^T\bm{f}^* - \bm{\psi}}} = 0,
\]
where $\bm{1}$ denotes the vector of ones and $\bm{\psi}_i = \psi(\bm{v}_i)$ is the entropy potential in terms of the nodal value of $\bm{u}_i$.  
\label{thm:dgsem}
\end{theorem}
\begin{proof}
Let $\bm{v}_i = \bm{v}(\bm{u}_i)$ denote the entropy variables, defined in terms of the nodal values $\bm{u}_i$.  We multiply (\ref{eq:dgsem}) by $\LRp{\bm{M}\bm{v}}^T$.  
Using the fact that $\bm{M}$ is diagonal, the time derivative term can be manipulated to 
\[
\bm{v}^T\bm{M}\pd{\bm{u}}{t} = \sum_{i=1}^{N+1} \bm{M}_{ii} \bm{v}_i^T \pd{\bm{u}_i}{t} = \sum_{i=1}^{N+1} \bm{M}_{ii} \pd{S_i}{t} = \bm{1}^T\bm{M}\pd{S_i}{t},
\]
where we have used continuity in time to arrive at the pointwise relation 
\[
\bm{v}^T \pd{\bm{u}}{t} = \pd{U(\bm{u})}{\bm{u}}^T \pd{\bm{u}}{t} = \pd{U(\bm{u})}{t}.
\]

The spatial derivative term can be manipulated using the summation by parts property $\bm{M}\bm{D} = \bm{S} = \bm{B}-\bm{S}^T$
\begin{align}
\sum_{i,j=1}^{N+1} \bm{v}_i^T 2 \bm{M}\bm{D}_{ij}\bm{f}_S(\bm{u}_i,\bm{u}_j) &= \sum_{i,j=1}^{N+1} \bm{v}_i^T (\bm{S} + \bm{B} - \bm{S}^T)_{ij}\bm{f}_S(\bm{u}_i,\bm{u}_j) \nonumber\\
&= \sum_{i,j=1}^{N+1} \bm{S}_{ij}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j) - \bm{S}_{ji}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j) + \bm{B}_{ij}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j). \label{eq:sbpentropy}
\end{align}
Rearranging indices and using symmetry of $\bm{f}_S(\bm{u}_i,\bm{u}_j)$ then yields that
\begin{align*}
\sum_{i,j=1}^{N+1} \bm{S}_{ij}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j) - \bm{S}_{ji}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j) &= \sum_{i,j=1}^{N+1} \bm{S}_{ij} \LRp{\bm{v}_i - \bm{v}_j}^T\bm{f}_S(\bm{u}_i,\bm{u}_j)\\
&= \sum_{i,j=1}^{N+1} \bm{S}_{ij} \LRp{\bm{\psi}_i-\bm{\psi}_j} = -\bm{1}^T \bm{S}\bm{\psi} = -\bm{1}^T\bm{B}\bm{\psi},
\end{align*}
where we have used that $\bm{S}\bm{1} = 0$ and the summation by parts property in the last line.

Using that $\bm{B}$ is diagonal, combining the latter term in (\ref{eq:sbpentropy}) with boundary terms involving the numerical flux yields
\[
\sum_{i,j = 1}^{N+1} \bm{B}_{ij}\bm{v}_i^T \bm{f}_S(\bm{u}_i,\bm{u}_j) + \bm{v}_i^T\bm{B}_{ij}\LRp{\bm{f}_i^* - \bm{f}(\bm{u}_i)} = \bm{1}^T\bm{B} \LRp{\bm{v}^T \bm{f}^*},
\]
where we have used the consistency of $\bm{f}_S(\bm{u},\bm{u}) = \bm{f}(\bm{u})$.  Combining everything results in a discrete statement of the conservation of entropy (\ref{eq:consentropy}).   
%Combining everything results in a discrete statement of the conservation of entropy (\ref{eq:consentropy}) 
%\[
%\bm{1}^T\LRp{\bm{M}\pd{U(\bm{u})}{t} + \bm{B}\LRp{\bm{v}^T\bm{f}^* - \bm{\psi}}} = 0.
%\]
\end{proof}

\begin{remark}
The proof of entropy conservation can be extended beyond a single domain by specifying appropriate simultaneous approximation terms (SBP-SAT).  This can be done by taking $\bm{f}_i^*$ to be the entropy conservative flux $f_S(u_L,u_R)$ at the interface between two elements \cite{carpenter2014entropy,gassner2017br1,chen2017entropy}.  Similarly, employing an entropy stable flux at element interfaces results in an entropy stable method which satisfies a global entropy inequality.  
\end{remark}

\section{Entropy conservative general DG methods}

In this section, we focus on constructing entropy conservative DG schemes with arbitrary bases and quadrature rules.  These include dense norm SBP operators and generalized SBP (GSBP) operators which do not contain boundary points, which introduce additional difficulties in constructing entropy conservative DG schemes.  These difficulties present themselves primarily in three areas:
\begin{enumerate}
\item It is unclear how to generalize the flux differencing approach to arbitrary choices of quadrature.  
\item The proof of entropy stability relies on a diagonal norm mass matrix, which commutes with pointwise multiplication by the (nodal values of) the entropy variables.  
\item Entropy stable formulations based on dense norm or GSBP operators require non-trivial SBP-SAT terms \cite{ranocha2017extended}.  These can be derived by hand for simple problems such as Burgers' equation, but can be difficult to determine for more complex systems of nonlinear PDEs.  
\end{enumerate}

We simultaneously address all three areas of difficulty by introducing a generalization of the entropy conservative DG-SEM scheme (\ref{eq:dgsem}) to arbitrary quadratures and choices of basis.  Additionally, we prove that the generalized scheme is entropy conservative (stable) on multiple elements in the following sections.

%We will get around these difficulties for dense-norm DG methods by satisfying three requirements for entropy stability
%\begin{itemize}
%\item $\bm{D}\bm{1} = 0$ for use of the flux formulation.  
%\item For some norm matrix $\bm{W}$, $\bm{W}\bm{D} = \bm{B}-\bm{D}^T\bm{W}$ where $\bm{B}$ is a suitably defined boundary operator.
%\item Contraction with the \textit{projection} of the entropy variables.  
%\end{itemize}

\subsection{A continuous interpretation of flux differencing}

We define the quadrature-based $L^2$ projection $\Pi_N$ as the $L^2$ projection where integrals are evaluated using some (inexact) quadrature.  We also define the discrete approximation space $V_h$, and assume that the quadrature is sufficiently accurate such that integration by parts holds for any two elements of $V_h$.  Finally, we assume that the discrete derivative (to be defined) maps from $V_h$ to $V_h$.
